---
title: "h5array Introduction"
author: "Paul Theodor Pyl"
date: "22 Jan 2016"
output: pdf_document
---

This vignette introduces the `h5array` R/Bioconductor package, which provides `array`- and `matrix`-like objects that store their data in HDF5 files on disk while being useable with the same interface as normal `array` and `matrix` objects.

The following code block illustrates basic usage of the package:

```{r, echo=FALSE, warning=FALSE}
suppressPackageStartupMessages(require(h5array))
suppressPackageStartupMessages(require(rhdf5))
```

The `h5arrayCreate` and `h5matrixCreate` functions are used to create `h5array` and `h5matrix` objects, these functions are only called once to create the dataset in the HDF5 file on disk.
```{r}
filename = tempfile()
x <- h5matrixCreate(
  filename,
  location = "/NumericMatrix",
  dim = c(1e5, 1e3), storage.mode = "double",
  chunk = c(1e3, 1e2))
x[1:23,] <- matrix(rnorm(23*1e3), nrow = 23)
```

We can user the `rhdf5::h5ls` function to inspect the HDF5-file that was created and see which datasets have been created within it:
```{r}
h5ls(filename)
```

The `h5matrixCreate` and `h5arrayCreate` function hand over any additional arguments to the internal `h5createDataset` function from the `rhdf5` package, therefore things like storage mode and chunking can be configured as well. If you are unfamiliar with storage modes and chunking it might help to have a look at the `rhdf5` vignette as well.

Let's create a string array in the same file:
```{r}
y <- h5arrayCreate(
  filename, #same file
  location = "/CharArray", #New location
  dim = c(3, 4, 5), #since we want 3d data we need to use the h5array class
  storage.mode = "character",
  size = 256 #to store strings in HDF5 files we need to specify the maximum length
  )
y[,,1] <- matrix(sample(letters, size = 3*4), nrow = 3)
```

Investigating the file again we can see the new dataset has been created.
```{r}
h5ls(filename)
```

We can also `print` the objects to get a descriptive representation of them:
```{r}
print(x)
print(y)
```

Adding dimension names is straightforward:
```{r}
dimnames(y) <- list(
  letters[1:3],
  letters[4:8],
  letters[9:14]
  )
print(y)
```

Dimension names are not stored in the HDF5 file by default but can be written there:
```{r}
writeDimnamesToFile(y)
h5ls(filename)
```

Note how new datasets to store the dimension names have been created.

Once the underlying datasets have been created the actual `h5array` or `h5matrix` objects can be removed (e.g. when closing the R session) and the data can be accessed usign the `h5array` and `h5matrix` functions:

```{r}
rm(x)
rm(y)
a <- h5matrix(filename, "/NumericMatrix")
b <- h5array(filename, "/CharArray")
print(a)
print(b)
```

We can load the dimnames from the file as well:
```{r}
b <- loadDimnamesFromFile(b)
print(b)
```